The role of advection versus diffusion in intraventricular transport

### 2025-02-03 paper/main.tex
add Introduction

\documentclass[fleqn]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage{lipsum}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{siunitx}
\usepackage{physics}
\usepackage{stmaryrd} % double brackets for jump operator
\usepackage{nameref}
\usepackage{booktabs}
\usepackage{makecell} % for table cells with line breaks
\usepackage{bm}
\usepackage{amsopn}
\usepackage[skip=10pt]{caption}
\usepackage{color}
\usepackage{xcolor,colortbl}
%\usepackage{soul}
\usepackage[nameinlink, capitalise]{cleveref}
\usepackage{graphicx}
\usepackage{graphbox} % for center-aligning figures vertically
\usepackage{tkz-euclide} % draw tikz pictures
\usepackage{subcaption}
\usepackage{tablefootnote} % for table footnotes
\usepackage{float} % figure placement

@@

% Bold faced math characters
\newcommand{\ff}{\mathbf{f}}
\newcommand{\nn}{\mathbf{n}}
\newcommand{\rr}{\mathbf{r}}
\newcommand{\uu}{\mathbf{u}}
\newcommand{\uuref}{\mathbf{u}_{\mathrm{ref}}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\ww}{\mathbf{w}}
\newcommand{\xx}{\bm{x}}
\newcommand{\JJ}{\mathbf{J}}
\newcommand{\VV}{\mathbf{V}}
\newcommand{\bsig}{\bm{\sigma}}
\newcommand{\bsigpar}{\hat{\bsig}_{\parallel}}
\newcommand{\bsigperp}{\hat{\bsig}_{\perp}}
\newcommand{\beps}{\bm{\varepsilon}}
\newcommand{\btau}{\bm{\tau}}

@@

\begin{abstract}
    This paper presents ...
    %Flow of cerebrospinal fluid in zebrafish brain ventricles is studied in this work. Motivated by experimental observations, we consider motile cilia and cardiac beating as the driving flow mechanisms. A mathematical model of the flow is developed and coupled with a transport model to study the transport of molecules inside the brain ventricles. The Stokes equations are used to model the flow, with the motile cilia acting as a force on the fluid through a stress boundary condition that allows slip at the wall. Transport is modeled with an advection-diffusion equation coupled to the velocity field obtained from solving the Stokes equations. The flow and transport models are discretized with finite element methods. For the Stokes equations we employ a non-conforming discretization based on elements of the Brezzi-Douglas-Marini family. The advection-diffusion equation is discretized with discontinuous Galerkin elements. We implement and solve the discretized equations numerically to simulate flow and transport in a zebrafish brain ventricles geometry obtained with medical imaging.
\end{abstract}
\begin{document}
\maketitle
% Introduction:
% 1. What are cilia and why are they important?
% 2. Previous studies of cilia flow
% 3. What problem do we want to solve
% 4. The model and work done in this paper
\section*{Introduction}
\lyng{HH: This needs work, have just jotted down some stuff to have references ready.}

Development of brains and sustaining healthy functioning of the brain relies on a myriad of regulatory processes. One of these processes is the flow of cerebrospinal fluid (CSF) in the brain ventricles. The brain ventricles are hollow structures inside the skull where CSF is produced and flows around, transporting nutrients, neurotransmitters and other molecules important for neural development.

The inner walls of the brain ventricles are lined with an epithelial cell layer. In some regions of the ventricles, motile cilia protrude the epithelial cell layer. Motile cilia are hair-like structures that beat in a metachronal wave pattern such as to generate a complicated flow pattern in the brain ventricles. Motile cilia are also found in other parts of the body where fluids are transport, such as the airways, liver and intestines. Regarding the function of motile cilia in the brain, it has previously been pointed out that further studies are needed for investigating their role~\cite{Eichele2020Cilia-drivenVentricle}, as well as the role of spatially organized CSF flow~\cite{Olstad2019CiliaryDevelopment}, in the development of the brain and its ventricular system.

Experiments~\cite{Olstad2019CiliaryDevelopment} have demonstrated that cilia induce a directional flow in the brain ventricles. In addition to the directional flow induced by motile cilia, a pulsatile component in the CSF flow in zebrafish ventricles has been observed~\cite{Olstad2019CiliaryDevelopment}. Cilia are also found to serve crucial roles in other organs than the brain. Motile cilia are essential for olfactory computations~\cite{Reiten2017Motile-Cilia-MediatedComputations}, and they are also fundamental to mucous transport in the trachea~\cite{Sleigh1988TheCilia}.

Early computational studies modeled ciliary transport of mucous in lungs with mathematics~\cite{Fulford1986Muco-ciliaryLung}. Propulsion of ciliated organisms has for some time been studied with mathematics~\cite{BLAKE1974MechanicsMotion}. According to Brennen~\emph{et al.}~\cite{Brennen1977FluidFlagella}, the study of cilia locomotion and fluid systems utilizing cilia sparked interest after a review on locomotion of protozoa microorganisms performed by Jahn~\emph{et al.}~\cite{Jahn1972LocomotionProtozoa}. Fine-scale interaction between single or several cilia and surrounding fluids has previously been analyzed with mathematics and numerics~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids, Cui2019NumericalMethod, Cui2022AFlow}, also specifically for cilia in brain ventricles of zebrafish embryos~\cite{Salman2022ComputationalEmbryo}. Quantification of bulk flow and effects on bulk transport by cilia arrays (or sheets) have not been studied as much, though by some~\cite{Ramirez-SanJuan2020Multi-scaleArrays}. Two-dimensional modeling of flow and transport mediated by motile cilia in zebrafish brain central canal has previously been carried out~\cite{Thouvenin2020OriginCanal}. There is, however, very limited research on three-dimensional modeling of cilia-mediated flow and transport in brain ventricles. There are a couple of studies indicating that the effect of cilia on CSF flow is local in the human brain~\cite{Siyahhan2014FlowVentricles, Yoshida2022EffectVentricles}. Zebrafish ventricles are much smaller than human ventricles, meaning cilia play a larger role in establishing a CSF flow pattern. We wish to investigate the mechanisms of ciliary and cardiac-beating flow components, and how they each contribute to transport of molecules inside the brain ventricles.

In this paper, we present protein photoconversion experiments in zebrafish embryo brain ventricles, as well as a computational model of cerebrospinal fluid flow and protein transport.

\section*{Methods}
\label{sec:methods}

>>>>

\documentclass[fleqn]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage{lipsum}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{siunitx}
\usepackage{physics}
\usepackage{stmaryrd} % double brackets for jump operator
\usepackage{nameref}
\usepackage{booktabs}
\usepackage{makecell} % for table cells with line breaks
\usepackage{bm}
\usepackage{amsopn}
\usepackage[skip=10pt]{caption}
\usepackage{color}
\usepackage{xcolor,colortbl}
\usepackage[nameinlink, capitalise]{cleveref}
\usepackage{graphicx}
\usepackage{graphbox} % for center-aligning figures vertically
\usepackage{tkz-euclide} % draw tikz pictures
\usepackage{subcaption}
\usepackage{tablefootnote} % for table footnotes
\usepackage{float} % figure placement

@@

% Bold faced math characters
\newcommand{\ff}{\mathbf{f}}
\newcommand{\nn}{\mathbf{n}}
\newcommand{\rr}{\mathbf{r}}
\newcommand{\uu}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\ww}{\mathbf{w}}
\newcommand{\xx}{\bm{x}}
\newcommand{\JJ}{\mathbf{J}}
\newcommand{\VV}{\mathbf{V}}
\newcommand{\bsig}{\bm{\sigma}}
\newcommand{\bsigpar}{\hat{\bsig}_{\parallel}}
\newcommand{\bsigperp}{\hat{\bsig}_{\perp}}
\newcommand{\beps}{\bm{\varepsilon}}
\newcommand{\btau}{\bm{\tau}}

@@

\begin{abstract}
    This paper presents ...
    %Flow of cerebrospinal fluid in zebrafish brain ventricles is studied in this work. Motivated by experimental observations, we consider motile cilia and cardiac beating as the driving flow mechanisms. A mathematical model of the flow is developed and coupled with a transport model to study the transport of molecules inside the brain ventricles. The Stokes equations are used to model the flow, with the motile cilia acting as a force on the fluid through a stress boundary condition that allows slip at the wall. Transport is modeled with an advection-diffusion equation coupled to the velocity field obtained from solving the Stokes equations. The flow and transport models are discretized with finite element methods. For the Stokes equations we employ a non-conforming discretization based on elements of the Brezzi-Douglas-Marini family. The advection-diffusion equation is discretized with discontinuous Galerkin elements. We implement and solve the discretized equations numerically to simulate flow and transport in a zebrafish brain ventricles geometry obtained with medical imaging.
\end{abstract}
\begin{document}
\maketitle
% Introduction:
% 1. What are cilia and why are they important?
% 2. Previous studies of cilia flow
% 3. What problem do we want to solve
% 4. The model and work done in this paper
\section*{Introduction}
Brain development and function depend on numerous regulatory processes, including interactions with cerebrospinal fluid (CSF).
CSF is a clear liquid produced by the choroid plexus, which is a secretory epithelium located within the brain ventricles.
Following production, CSF flows through the ventricular system and subarachnoid space, where it regulates brain development and
homeostasis by transporting growth factors, nutrients, neurotransmitters, waste products and other molecules and metabolites.

To achieve its function, CSF needs to circulate throughout the brain. In fact, CSF disturbances can univocally lead to brain disorders,
such as hydrocephalus where accumulation of CSF requires emergency surgical treatment.
CSF movement occurs through multiple physiological processes, including the action of motile cilia on the ventricular wall,
CSF secretion by the choroid plexus and pressure gradients generated by the cardiac cycle, respiration, and bodily movement.
Motile cilia are microscopic hair-like structures that beat in a metachronal wave pattern to generate flow pattern. Due to their microscopic size, ranging from 6 to 15 microns,
they contribute to fluid flow primarily at the surface of the brain ventricles.
In contrast, pressure gradients generate fluid flow further from the ventricular walls and along the perivascular space surrounding the major blood vessels.

To date, it remains poorly understood how cilia contribute to the overall fluid flow and solute transport in the brain.
These questions have been especially challenging to tackle due to technical limitations.
Notably, cilia are very small and beat at high frequencies ranging from 10 to 30 Hz, while other fluid dynamic processes in the brain are slower and span larger three-dimensional volumes.
Most studies until now have analyzed one process at a time, especially in mammals.
For instance, multiple studies have convincingly shown that cilia generate complex flow patterns at the surface of the brain ventricles.
However, since all these studies have been performed on brain explants, they failed to identify the role of pressure gradients associated with respiration and heart pulsations in CSF transport.
Similarly, while MRI technology can measure CSF movement in humans, it cannot detect the fast movements of cilia or understand their impact in the large human brain.
Consequently, we lack a comprehensive understanding of how cilia contribute to CSF movement and solute transport, and whether cilia dysfunction will alter CSF distribution and lead to brain disorders.
This is particularly important since ventricular enlargement or hydrocephalus is commonly observed in animal models with motile cilia deficiency, potentially affecting brain activity and behavior.

The technical limitations of studying cilia-mediated flow and transport opens the possibility of computational investigations. Cilia have been studied previously with mathematical modeling,
with numerous studies on the interaction between single or several cilia and surrounding fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids, Cui2019NumericalMethod, Cui2022AFlow}.
Others consider applications such as ciliary transport of mucous in airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion, Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
coral reefs~\cite{Pacherres2022CiliaryProduction}, and the brain~\cite{Siyahhan2014FlowVentricles, Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.

In this study, we combined computational and experimental approaches to investigate the role of motile cilia in CSF flow generation and solute transport.
First, we computed flow patterns and intraventricular transport in silico. To this end, we used the geometry of the zebrafish larval brain ventricles,
for which we have detailed knowledge of cilia properties and CSF movement velocities and frequencies.
Notably, this early developmental stage allows us to simplify our model by omitting CSF secretion, which starts at later stages, while still being well conserved with mammals.
Next, we validated our computation model with experimental data by monitoring the dynamics of a photoconvertible fluorescent protein expressed within the brain ventricles.
Our results identified that diffusion plays a major role in the transport of small solutes, whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia and the ventricular geometry greatly impact solute distribution.
Altogether, our work presents a computational framework and new concepts of how cilia collaborate with other factors to distribute molecules and particles within the brain ventricles.

\section*{Methods}
